Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-10 linear_adia.tex
appendix.tex
corrected two typos (eq 67 and eq 46)
updated iso discussion in appendix

\subsection{Thin-disk approximation}\label{analytic_relax}
Our final approximation apply to thin disks for 
which $\epsilon\ll 1$. Then we may write $\rho/\rho_0\simeq
\exp{(-\zhat^2/2)}$. The governing equation,
Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  0 = \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    \epsilon q\hat{k}\right), 
\end{align}
and $\chi =
\left(1-\ii\hat{\sigma}\beta\right)/\left(1-\ii\hat{\sigma}\beta\gamma\right)
$ in terms of dimensionless variables.

@@

In Appendix \ref{iso_discuss}, we give an analytical description of the
linear problem with $\beta=0$. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability (there must be vertical shear)
and an upper bound to the instability growth rate (it is limited by
the maximum vertical shear rate in the disk). We also develop analytic
solutions with and without vertical shear in the thin disk limit,
including explicit expressions for wave frequencies and growth rates. 
The methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections.

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI mode for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. In Appendix \ref {iso_stable} we discuss
stable solutions without vertical shear which yield
$\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(Eq. \ref{iso_stable_disp}).  We are thus motivated to seek solutions
to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

>>>>

\subsection{Thin-disk approximation}\label{analytic_relax}
Our final approximation apply to thin disks for 
which $\epsilon\ll 1$. Then we may write $\rho/\rho_0\simeq
\exp{(-\zhat^2/2)}$. The governing equation,
Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii \epsilon q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    \epsilon q\hat{k}\right), 
\end{align}
and $\chi =
\left(1-\ii\hat{\sigma}\beta\right)/\left(1-\ii\hat{\sigma}\beta\gamma\right)
$ in terms of dimensionless variables.

@@

In Appendix \ref{iso_discuss}, we give an analytical discussion of the
linear problem with $\beta=0$. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability (there must be vertical shear)
and an upper bound to the instability growth rate (it is limited by
the maximum vertical shear rate in the disk). We also develop analytic
solutions in the thin disk limit. The methods used in Appendix
\ref{iso_discuss} and results obtained therein motivates the solution
procedure employed in the following sections.

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI mode for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3 \notag\\
    &+ 2\beta\gamma(\hat{\omega}\khat)^4.  
\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. In Appendix \ref {iso_stable} we discuss
stable solutions without vertical shear which yield
$\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(Eq. \ref{iso_stable_disp}).  We are thus motivated to seek solutions
to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

### 2015-03-10 numerical.tex
updated iso discussion in appendix

Fig. \ref{iso_eigen_kx} compares the numerically-obtained eigenvalues
for the fundamental mode and that calculated from 
Eq. \ref{sig2_iso} (with $L=1$). There is good agreement for 
$\khat\gtrsim10$. The match worsens for smaller $\khat$ because the 
the low-frequency approximation used to obtain Eq. \ref{sig2_iso} 
is no longer valid, as $|\omega|$ is not small compared to $\Omega_k$
for $\khat\lesssim 10$.

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in Appendix \ref{iso_poly}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}. The fundamental mode
consists of zero nodes in $\delta v_z$.

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental mode has the
smallest $|\sigma|$, for which the eigenvalue calculated from
Eq. \ref{sig2_iso} with $L=1$ and the numerically-obtained value are   
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

>>>>

Fig. \ref{iso_eigen_kx} compares the numerically-obtained eigenvalues
for the fundamental mode and that calculated from
Eq. \ref{simple_growth} (with $L=1$). There is good agreement for  
$\khat\gtrsim10$. The match worsens for decreasing $\khat$ due to the
neglect of global radial structure in the analysis.

% the low-frequency approximation used to obtain Eq. \ref{sig2_iso} 
% is no longer valid, as $|\omega|$ is not small compared to $\Omega_k$
% for $\khat\lesssim 10$.

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in Appendix \ref{iso_explicit}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}. The fundamental mode
consists of zero nodes in $\delta v_z$.

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental mode has the
smallest $|\sigma|$, for which the eigenvalue calculated from
Eq. \ref{simple_growth2} with $L=1$ and the numerically-obtained value are   
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2605)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{simple_growth2})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{simple_growth2}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.
